### 2022-02-17 chapter1/chapter.tex
first draft of the whole thing! :D

%!TEX root = ../thesis.tex
\chapter{The Existing Literature \& Mathematical Preliminaries \label{ch:math}} 
% intro to intro
% contextual background
	% Fluid queues 
		% definition
		% applications 
		% existing results 
	% Fluid-fluid queues
		% definition 
		% basic analysis in terms of generators
		% the need for approximation and how the old work doesn't fit here
	% The DG method
		% pertition into cells
		% project on to a basis
		% problems: oscillations
		% soutions and why they dont fit
		% observation about constant-basis methods
		% this thesis asks the question: motivated by the constant basis being a probability model, can we dream up a more accurate probability model to approximate a fluid queue?
	% On the structure of the constant basis/uniformisation method (QBD)
	% least variable is PH, so what about MEs?
% structure of the thesis
	% rest of chapter 1 gives mathematical preliminaries
	% chapter 2 explains the DG method, problems/oscillations, slope limiting, and appllication to SFFMs
	% inspired by the structure of the order-1 scheme, and to solve the negativity problem, chapter 3 introduces a new approximation method; The QBD-RAP. In this chapter we derive the intuitively of the model and explain it's dynamics.
	% we then move to convergence, with chapter 4 proving a convergence of the QBD-RAP up to the first hitting time of the fluid queue on the edges of an interval. technical results and extensions are left to the appendix: certain properties of closing vectors, convergence without ephemeral phases, and certain matrix algebraic manipulations. 
	% chapter 5 stitches together the results of chapter 4 to prove a global result about convergence
	% chapter 6 investigate, numerically, the performance of the DG, DG with limiter, uniformisation and QBD-RAP schemes. 
	% chapter 7  makes concluding remarks.
% mathematical preliminaries 
	% CTMCs
	% fluid queues
	% fluid-fluid queues and operator-analytic expressions (all of them)
		% the equations of bo2014
		% meets the partition of the DG/QBD-RAP, and reconstructing the Fil sets 
    % why we cant solve the equations, what we need to approximate to solve it, i.e. B, R, then D, Psi
	% projections
	% phase-type distributions 
		% definition
		% least variable property
	% matrix exponentials
		% definition 
		% properties 
		% verification that parameters give an ME
		% CMEs
	% QBD-RAPs, then QBDs as a special case
		% orbit processes, their interpretation for PH and how they differ for MEs
	% convergence theorems 
		% DCT 
		% Continuity of Laplace transforms

@@

A stochastic process is a sequence, \(\{X(t)\}_{t\in\mathcal T}\), of random variables or random vectors indexed by some index set \(\mathcal T\). In this thesis in the case that the index set is \(\mathcal T=\{0,1,2,...\}\) we say that the process \(\{X(n)\}_{n\in \mathcal T}\) is a discrete-time process, and we will typically use the dummy variable \(n\). When the index set is \(\mathcal T=[0,\infty)\) we say write \(\{X(t)\}_{t\in\mathcal T}\) is a continuous-time process, and we will typically use the dummy variable \(t\). We may omit the index set and write \(\{X(t)\}\) in place of \(\{X(t)\}_{t\in\mathcal T}\) when it is not explicitly needed, or we may write \(\{X(t)\}_{t\geq t_0}\) to mean \(\{X(t)\}_{t\in[t_0,\infty)}\). The \emph{sample space} of \(X(t)\) is the set of possible values that \(X(t)\) can take at any \(t\in\mathcal T\).

The \emph{initial distribution}, \(\mu\), of a stochastic process is the distribution of \(X(0)\). More generally, for a random variable, \(Z\), we write \(Z\sim \nu\) when \(Z\) has the distribution given by the probability measure \(\nu\). For the probability that \(X(t)\) lies in some set \(E\subset \mathcal S\) we write \(\mathbb P(X(t)\in E \mid X(0)\sim \mu)\). When \(\mu\) assigns probability 1 to a single point, \(x\in\calS\), say, we write \(\mathbb P(X(t)\in E \mid X(0)=x).\) A stochastic process is \emph{stationary} if
\begin{align}\label{eqn: stationary}
	&\nonumber\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	\\&= \mathbb P(X(t_n+t)=i,X(t_{n-1}+t)\in E_{n-1},...,X(t_0+t)\in E_0),
\end{align}
for any \(t,t_0,...,t_n\in\mathcal T\) and any \(E_0,...,E_n\subseteq \mathcal S\).
A random variable, \(\tau\), is a \emph{stopping time} for the stochastic process \(\{X(t)\}_{t\in\mathcal T}\) if \(\tau\in\sigma(X(s), s\leq t),\) where \(\sigma(X(s), s\leq t)\) denotes the \(\sigma\)-algebra generated by \(\{X(s), s\leq t\}\).

We call \(\{X(n)\}_{n\in\{0,1,2,\dots\}}\) a \emph{discrete-time Markov chain} if \(\calS\) is countable
\begin{align}\label{eqn: markov property}
	\nonumber &\mathbb P(X(n+1)=j \mid X(n)=i,X(n-1)=i_{n-1},...,X(0)=i_0) 
	\\&= \mathbb P(X(n+1)=j \mid X(n)=i) 
\end{align}
for all \(n\in\{0,1,2,...\}\) and \(i,i_0,...,i_{n-1},j\in\calS,\) and refer to (\ref{eqn: markov property}) as the Markov property. The process \(\{X(n)\}_{n\in\{0,1,2,...\}}\) is said to be time-homogeneous if 
\(\mathbb P(X(n+1)=j \mid X(n)=i)\) does not depend on \(n\). The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(n)\}_{n\in\{0,1,2,...\}}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong markov property}
	\mathbb P(X(\tau+1)=j \mid \sigma(X(0),X(1),...,X(\tau))) = \mathbb P(X(\tau+1)=j \mid X(\tau)).
\end{align}

We call \(\{X(t)\}_{t\geq 0}\) a \emph{continuous-time Markov chain} if \(\calS\) is countable and for all \(t_{n+1} > t_n > t_{n-1} > ... > t_0 > 0\) and \(i,i_0,...,i_{n-1},j\in\calS,\)
\begin{align}\label{eqn: markov property cmtc}
	\mathbb P(X(t_{n+1})=j \mid X(n)=i, X(n-1)=i_{n-1},...,X(0)=i_0) = \mathbb P(X(t_{n+1})=j \mid X(t_n)=i).
\end{align}
The process \(\{X(t)\}_{t\geq 0}\) is said to be time-homogeneous if 
\(\mathbb P(X(t+s)=j \mid X(s)=i)\) does not depend on \(s\). The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(t)\}_{t\geq 0}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong markov property ctmc}
	\mathbb P(X(\tau + t)=j \mid \sigma(X(s),s\leq \tau)) = \mathbb P(X(\tau+t)=j \mid X(\tau)).
\end{align}

Most generally, we call \(\{X(t)\}_{t\geq 0}\) a \emph{Markov process} if for all \(t_n > t_{n-1} > ... > t_0 > 0,\, t_k\in\mathcal T,\, k=0,...,n+1,\) and \(E_0,...,E_n\subseteq \mathcal S\)
\begin{align}\label{eqn: markov process}
	\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	= \mathbb P(X(t_n)\in E_n\mid X(t_{n-1})),
\end{align}
and is time homogeneous if \(\mathbb P(X(t+s)\in E_n\mid X(s))\) does not depend on \(s\).

The Chapman-Kolmogorov equations state that 
\begin{align}\label{eqn: CK}
	&\nonumber \mathbb P(X(t+s)\in E\mid X(0)\sim \mu) 
	\\&= \int_{x\in\calS}\mathbb P(X(t+s)\in E\mid X(t)=x)\mathbb P(X(t)\in \wrt x\mid X(0)\sim \mu).
\end{align}
% Markov processes, time-homogeneity, strong markov prperty, partitioning and LOTP

\section{Some semigroup theory}
The evolution of Markov processes can be described by an \emph{operator semigroup}. Furthermore, semigroups arise in other contexts related to the analysis of fluid queues and fluid-fluid queues. Semigroups (and therefore Markov processes) can be characterised by their \emph{infinitesimal generator}, or generator for short. One of the aims of this thesis is to approximate the generator of a type of Markov process called a fluid queue, which we introduce more formally in Section~\ref{sec: fqs}. Here we very briefly introduce semigroups and their \emph{infinitesimal generator}.

\subsubsection{Operators}
Let \(L\) be a Banach space (a vector space with a metric and where Cauchy sequences converge to a point within the space). An operator is a mapping which takes elements of a vector space and sends them to elements of a vector space (not necessarily the same space). That is, let \(V,W\) be normed vector spaces, then \(\mathbb B:\mathcal D(\mathbb B)\subset V\to W\) is an operator. The set \(\mathcal D(\mathbb B)\) is the domain of the operator, and is not necessarily all of \(L\), when specification of the domain is unnecessary we sometimes write \(\mathbb B:V\to W\). Let \(c\in \mathbb R\) and \(v\in V\), \(w\in W\). Then \(\mathbb B\) is a linear operator if \(\mathbb B(cv) = c(\mathbb Bv)\) and \(\mathbb B(v+w)=\mathbb Bv+\mathbb Bw\). We define the norm of an operator as \(||\mathbb B||=\sup_{v\in V}\cfrac{||\mathbb Bv||_W}{||v||_V}\), where \(||\cdot||_V\) and \(||\cdot||_W\) are the norms attached to \(V\) and \(W\) respectively.  An operator is said to be bounded if \(||\mathbb B||<M\) for some \(\infty > M\in \mathbb R\), for ever \(v\in V\).

@@

\subsubsection{Semigroups}  A family of linear operators is a collection of indexed linear operators \(\{\mathbb B(i):i\in I\}\) where \(I\) is some index set. That is, for each \(i\in I\), \(\mathbb B(i)\) is a linear operator. A family of operators \(\{T(t):L\to L:t\geq 0\}\) is said to have the \emph{semigroup} property if \(\mathbb T(t+s)=\mathbb T(t)T(s)\). If, furthermore, \(\mathbb T(0)=I\), the identity operator, then the family \(\{\mathbb T(t):t\geq 0\}\) is known as a semigroup. If \(\mathbb T(t)f\to f\) as \(t\to 0^+\), then \(\{\mathbb T(t):L\to L:t\geq 0\}\) is said to be strongly continuous. If the \(||\mathbb T(t)||\leq 1\) for all \(t\geq 0\) then \(\{\mathbb T(t):L\to L:t\geq 0\}\) is known as a contraction semigroup.

For example, for a bounded linear operator \(\mathbb B\), define the operator exponential as \(e^{\mathbb Bt}:=\sum_{k=0}^\infty\cfrac{1}{k!}t^k\mathbb B^k\), then it can be shown that \(e^{\mathbb B(t+s)}=e^{\mathbb Bt}e^{\mathbb Bs}\), and from the definition \(e^{\mathbb B0}=I\), hence \(\{e^{\mathbb Bt}:t\geq 0\}\) is a semigroup. As a more concrete example, let \(\mathbb B=\cfrac{d}{dx}:C^\omega \to C^\omega(\mathbb R)\), where \(C^\omega(\mathbb R)\) is the class of analytic functions on the whole of \(\mathbb R\). On \(C^\omega(\mathbb R)\), \(\cfrac{\wrt}{\wrt x}\) is bounded. Then \(e^{\frac{\wrt}{\wrt x} t } f(x) = \sum_{k=0}^\infty \cfrac{1}{k!}t^k\cfrac{\wrt^k}{\wrt x^k}f(x)=f(x-t)\), and the sum converges since \(f\in C^\omega(\mathbb R)\); the series representation is just the Taylor series of \(f\) about the point \(x\).

\subsubsection{Infinitesimal generators} The \emph{infinitesimal generator} (or just \emph{generator} for short) of a semigroup \(\mathbb T(t):L\to L\) is the linear operator defined by 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(T(t)f-I),\]
and the domain, \(\mathcal D(\mathbb B)\) is the subspace of \(L\) for which this limit exists. Note that an essential part of the definition of the generator is its domain. For this reason it is common to denote the generator as the pair \((\mathbb B,\mathcal D(\mathbb B))\). Here we will denote the generator as just \(\mathbb B\) and explicitly state its domain where relevant.

To determine the generator \(\mathbb B\) we can proceed by either differentiation of the semigroup, 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(T(t)f-I),\]
or integration;
\[\int_{t=0}^\infty e^{-st}T(t)\wrt t = (sI-\mathbb B)^{-1}=:R_s.\]
The operator \(R_s\) is known as the resolvent.

% \section{Markov processes and semigroups}
% Semigroups arise naturally in Markov processes. Let \(\{X(t):t\geq 0\}\) be a Markov process with state space \(\mathcal S\). The Markov property states that for \(s,t\geq 0\), \(X(t+s)\) is independent of \(X(u),\, 0\leq u<t\) given \(X(t)\). That is, \(P(X(t+s)\in\mathcal A \mid X(t), X(u), 0\leq u<t)=P(X(t+s)\in\mathcal A \mid X(t))\). If \(P(X(t+s)\in\mathcal A \mid X(t))=P(X(s)\in\mathcal A \mid X(0))\), that is the distributions are invariant under translation by \(t\), then \(X(t)\) is said to be time-homogeneous. Where necessary, we will use the notation \(X_\mu(t)\) to denote the Markov process with initial distribution \(\mu\), i.e.~\(P(X(0)\in\cdot)=\mu(\cdot)\), and \(X_x(t)\) to be the Markov process with initial distribution \(\delta(x)\). Markov processes can be characterised by \emph{probability kernels}.

% Let \((U,\mathcal U)\) and \((V,\mathcal V)\) be two measure spaces, then a probability kernel \(\nu\), is a mapping \(\nu:U\times \mathcal V\to \mathbb R_+\) such that \(\nu(u,\mathcal A)\) is \(\mathcal U\)-measurable in \(u\in U\) for fixed \(\mathcal A\in \mathcal V\), and a probability measure in \(\mathcal A\in \mathcal V\) for fixed \(u\in U\). We say that \(\nu\) is a probability kernel from \(U\) to \(V\).

% With a probability kernel comes associated integral operators. For a measure \(\mu\) on \((U,\mathcal U)\), then 
% \[\mu \nu (\mathcal A)=\int_{U} \mu(\wrt x) \nu(x,\mathcal A)\]
% is also a measure on \((U,\mathcal U)\). For a function \(f:V\to \mathbb R\) then 
% \[\nu f (x)=\int_{V}  \nu(x,\wrt y)f(y).\] 
% We can also compose two transition kernels. Let \(\nu\) be a transition kernel from \(U\) to \(V\) and \(\eta\) be a transition kernel from \(V\) to \(W\). Then \(\nu\eta\) is also a probability kernel from \(U\) to \(W\);
% \[\nu\eta(x,\mathcal A) = \int_{V}\nu(x,\wrt y)\eta(y,\mathcal A).\]

% Let \(\mu_{t}(\cdot)=P(X(t)\in \cdot)\) for \(t\geq 0\) be a one-parameter family of measures on \((\mathcal S,\mathcal F)\), and \(\nu_{t}(x,\cdot)=P(X(t)\in\cdot \mid X(0)=x)\) for \(t\geq 0\) be a one-parameter family of probability kernels from \(\mathcal S\) to \(\mathcal S\). Then \(\mu_t\) and \(\nu_{t}\) characterise the time-homogeneous Markov process \(X(t)\). The probability kernel \(\nu_t\) is referred to as a transition kernel. The kernels \(\nu_t\) form a semigroup since, using the law of total probability and time-homogeneity, 
% \begin{align*}
% \nu_{t+s}(x,\mathcal A)&=P(X(t+s)\in\mathcal A\mid X(0)=x)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(t+s)\in\mathcal A\mid X(t)=y)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(s)\in\mathcal A\mid X(0)=y)
% \\&=\int_{y\in\mathcal S}\nu_t(x,\wrt y)\nu_s(y,\mathcal A)=\nu_t\nu_s(x,\mathcal A).
% \end{align*}

>>>>

%!TEX root = ../thesis.tex
\chapter{The existing literature \& mathematical preliminaries\label{ch:math}} 
% intro to intro
% contextual background
	% Fluid queues 
		% definition
		% applications 
		% existing results 
	% Fluid-fluid queues
		% definition 
		% basic analysis in terms of generators
		% the need for approximation and how the old work doesn't fit here
	% The DG method
		% pertition into cells
		% project on to a basis
		% problems: oscillations
		% soutions and why they dont fit
		% observation about constant-basis methods
		% this thesis asks the question: motivated by the constant basis being a probability model, can we dream up a more accurate probability model to approximate a fluid queue?
	% On the structure of the constant basis/uniformisation method (QBD)
	% least variable is PH, so what about MEs?
% structure of the thesis
	% rest of chapter 1 gives mathematical preliminaries
	% chapter 2 explains the DG method, problems/oscillations, slope limiting, and appllication to SFFMs
	% inspired by the structure of the order-1 scheme, and to solve the negativity problem, chapter 3 introduces a new approximation method; The QBD-RAP. In this chapter we derive the intuitively of the model and explain it's dynamics.
	% we then move to convergence, with chapter 4 proving a convergence of the QBD-RAP up to the first hitting time of the fluid queue on the edges of an interval. technical results and extensions are left to the appendix: certain properties of closing vectors, convergence without ephemeral phases, and certain matrix algebraic manipulations. 
	% chapter 5 stitches together the results of chapter 4 to prove a global result about convergence
	% chapter 6 investigate, numerically, the performance of the DG, DG with limiter, uniformisation and QBD-RAP schemes. 
	% chapter 7  makes concluding remarks.
% mathematical preliminaries 
	% CTMCs
	% fluid queues
	% fluid-fluid queues and operator-analytic expressions (all of them)
		% the equations of bo2014
		% meets the partition of the DG/QBD-RAP, and reconstructing the Fil sets 
    % why we cant solve the equations, what we need to approximate to solve it, i.e. B, R, then D, Psi
	% projections
	% phase-type distributions 
		% definition
		% least variable property
	% matrix exponentials
		% definition 
		% properties 
		% verification that parameters give an ME
		% CMEs
	% QBD-RAPs, then QBDs as a special case
		% orbit processes, their interpretation for PH and how they differ for MEs
	% convergence theorems 
		% DCT 
		% Continuity of Laplace transforms

@@

A stochastic process is a sequence, \(\{X(t)\}_{t\in\mathcal T}\), of random variables (or random vectors) indexed by some index set \(\mathcal T\). In this thesis, in the case that the index set is \(\mathcal T=\{0,1,2,...\}\) we say that the process \(\{X(n)\}_{n\in \mathcal T}\) is a discrete-time process, and we will typically use the dummy variable \(n\) for the `time' index. When the index set is \(\mathcal T=[0,\infty)\) we say \(\{X(t)\}_{t\in\mathcal T}\) is a continuous-time process, and we will typically use the dummy variable \(t\) for the `time' index. We may omit the index set and write \(\{X(t)\}\) in place of \(\{X(t)\}_{t\in\mathcal T}\) when it is not explicitly needed, or we may write \(\{X(t)\}_{t\geq t_0}\) to mean \(\{X(t)\}_{t\in[t_0,\infty)}\). The \emph{sample space} of \(X(t)\) is the set of possible values that \(X(t)\) can take at any time \(t\in\mathcal T\).

The \emph{initial distribution}, \(\mu\), of a stochastic process is the distribution of \(X(0)\). More generally, for a random variable, \(Z\), we write \(Z\sim \nu\) when \(Z\) has the distribution given by the probability measure \(\nu\). For the probability that \(X(t)\) lies in some set \(E\subset \mathcal S\) given \(X(0)\sim \mu\), we write \(\mathbb P(X(t)\in E \mid X(0)\sim \mu)\). When \(\mu\) assigns probability 1 to a single point, \(x\in\calS\), say, we write \(\mathbb P(X(t)\in E \mid X(0)=x).\)

A stochastic process is \emph{stationary} if
\begin{align}\label{eqn: stationary}
	&\nonumber\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	\\&= \mathbb P(X(t_n+t)=i,X(t_{n-1}+t)\in E_{n-1},...,X(t_0+t)\in E_0),
\end{align}
for any \(t,t_0,...,t_n\in\mathcal T\) and any \(E_0,...,E_n\subseteq \mathcal S\), i.e.~the joint distribution~(\ref{eqn: stationary}) does not depend on \(t\).
A random variable, \(\tau\), is a \emph{stopping time} for the stochastic process \(\{X(t)\}_{t\in\mathcal T}\) if \(\tau\in\sigma(X(s), s\leq t),\) where \(\sigma(X(s), s\leq t)\) denotes the \(\sigma\)-algebra generated by \(\{X(s), s\leq t\}\).

We call \(\{X(n)\}_{n\in\{0,1,2,\dots\}}\) a \emph{discrete-time Markov chain} if \(\calS\) is countable and
\begin{align}\label{eqn: Markov property}
	\nonumber &\mathbb P(X(n+1)=j \mid X(n)=i,X(n-1)=i_{n-1},...,X(0)=i_0) 
	\\&= \mathbb P(X(n+1)=j \mid X(n)=i) 
\end{align}
for all \(n\in\{0,1,2,...\}\) and \(i,i_0,...,i_{n-1},j\in\calS,\) and refer to (\ref{eqn: Markov property}) as the Markov property. The process \(\{X(n)\}_{n\in\{0,1,2,...\}}\) is said to be time-homogeneous if 
\(\mathbb P(X(n+1)=j \mid X(n)=i)\) does not depend on \(n\). The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(n)\}_{n\in\{0,1,2,...\}}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong Markov property}
	\mathbb P(X(\tau+1)=j \mid \sigma(X(0),X(1),...,X(\tau))) = \mathbb P(X(\tau+1)=j \mid X(\tau)).
\end{align}

We call \(\{X(t)\}_{t\geq 0}\) a \emph{continuous-time Markov chain} if \(\calS\) is countable and for all \(t_{n+1} > t_n > t_{n-1} > ... > t_0 > 0\) and \(i,i_0,...,i_{n-1},j\in\calS,\)
\begin{align}\label{eqn: Markov property cmtc}
	\mathbb P(X(t_{n+1})=j \mid X(n)=i, X(n-1)=i_{n-1},...,X(0)=i_0) = \mathbb P(X(t_{n+1})=j \mid X(t_n)=i),
\end{align}
and refer to (\ref{eqn: Markov property cmtc}) as the \emph{Markov property}.
The process \(\{X(t)\}_{t\geq 0}\) is said to be time-homogeneous if 
\(\mathbb P(X(t+s)=j \mid X(s)=i)\) does not depend on \(s\). The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(t)\}_{t\geq 0}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong Markov property ctmc}
	\mathbb P(X(\tau + t)=j \mid \sigma(X(s),s\leq \tau)) = \mathbb P(X(\tau+t)=j \mid X(\tau)).
\end{align}

Most generally, we call \(\{X(t)\}_{t\geq 0}\) a \emph{Markov process} if for all \(t_n > t_{n-1} > ... > t_0 > 0,\, t_k\in\mathcal T,\, k=0,...,n+1,\) and \(E_0,...,E_n\subseteq \mathcal S\)
\begin{align}\label{eqn: Markov process}
	\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	= \mathbb P(X(t_n)\in E_n\mid X(t_{n-1})),
\end{align}
and is time homogeneous if \(\mathbb P(X(t+s)\in E_n\mid X(s))\) does not depend on \(s\).

The Chapman-Kolmogorov equations state that 
\begin{align}\label{eqn: CK}
	&\nonumber \mathbb P(X(t+s)\in E\mid X(0)\sim \mu) 
	\\&= \int_{x\in\calS}\mathbb P(X(t+s)\in E\mid X(t)=x)\mathbb P(X(t)\in \wrt x\mid X(0)\sim \mu).
\end{align}
% Markov processes, time-homogeneity, strong Markov prperty, partitioning and LOTP

\section{Some semigroup theory}
The evolution of Markov processes can be described by an \emph{operator semigroup}. Semigroups also arise in other contexts related to the analysis of fluid queues and fluid-fluid queues (see, for example, Sections~\ref{sec: ffq intro} and~\ref{sec: transient ffq intro}). Semigroups (and therefore Markov processes) can be characterised by their \emph{infinitesimal generator}, or generator for short. One of the aims of this thesis is to approximate the generator of a fluid queue. Here we very briefly introduce semigroups and their \emph{infinitesimal generators}.

\subsubsection{Operators}
Let \(L\) be a Banach space (a vector space with a metric and where Cauchy sequences converge to a point within the space). An operator is a mapping which takes elements of a vector space and sends them to elements of a vector space (not necessarily the same space). That is, let \(V\) and \(W\) be normed vector spaces, then \(\mathbb B:\mathcal D(\mathbb B)\subset V\to W\) is an operator. The set \(\mathcal D(\mathbb B)\) is the domain of the operator, and is not necessarily all of \(V\). When specification of the domain is unnecessary we sometimes write \(\mathbb B:V\to W\). Let \(c\in \mathbb R\) and \(v\in V\), \(w\in W\). Then \(\mathbb B\) is a \emph{linear} operator if \(\mathbb B(cv) = c(\mathbb Bv)\) and \(\mathbb B(v+w)=\mathbb Bv+\mathbb Bw\). We define the norm of an operator as \(||\mathbb B||=\sup_{v\in V}\cfrac{||\mathbb Bv||_W}{||v||_V}\), where \(||\cdot||_V\) and \(||\cdot||_W\) are the norms attached to \(V\) and \(W\), respectively.  An operator is said to be bounded if \(||\mathbb B||<M\) for some \(M<\infty, M\in \mathbb R\), for every \(v\in V\).

@@

\subsubsection{Semigroups}  A family of linear operators is a collection of indexed linear operators \(\{\mathbb B(i)\}_{i\in I}\) where \(I\) is some index set. That is, for each \(i\in I\), \(\mathbb B(i)\) is a linear operator. A family of operators \(\{\mathbb T(t)\}_{t\geq 0}\), \(\mathbb T(t):L\to L\), is said to have the \emph{semigroup} property if \(\mathbb T(t+s)=\mathbb T(t)\mathbb T(s)\). If \(\mathbb T(0)=I\), the identity operator, then the family \(\{\mathbb T(t)_{t\geq 0}\) is known as a semigroup. If \(\mathbb T(t)f\to f\) as \(t\to 0^+\), then \(\{\mathbb T(t)\}_{t\geq 0}\) is said to be strongly continuous. If \(||\mathbb T(t)||\leq 1\) for all \(t\geq 0\) then \(\{\mathbb T(t)\}_{t\geq 0}\) is known as a contraction semigroup.

For example, for a bounded linear operator \(\mathbb B\), define the operator exponential as \(e^{\mathbb Bt}:=\sum_{k=0}^\infty\cfrac{1}{k!}t^k\mathbb B^k\), then it can be shown that \(e^{\mathbb B(t+s)}=e^{\mathbb Bt}e^{\mathbb Bs}\), and from the definition \(e^{\mathbb B0}=I\), hence \(\{e^{\mathbb Bt}:t\geq 0\}\) is a semigroup. As another example, let \(\mathbb B=\cfrac{d}{dx}:C^\omega \to C^\omega(\mathbb R)\), where \(C^\omega(\mathbb R)\) is the class of analytic functions on the whole of \(\mathbb R\). Then \(\sum_{k=0}^\infty \cfrac{1}{k!}t^k\cfrac{\wrt^k}{\wrt x^k}f(x)=f(x-t)\), and the sum converges since \(f\in C^\omega(\mathbb R)\); the series representation is just the Taylor series of \(f\) about the point \(x\).

\subsubsection{Infinitesimal generators} The \emph{infinitesimal generator} (or just \emph{generator} for short) of a semigroup \(\{\mathbb T(t)\}\) is the linear operator defined by 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(\mathbb T(t)f-I),\]
and the domain, \(\mathcal D(\mathbb B)\), is the subspace of \(L\) for which this limit exists. Note that an essential part of the definition of the generator is its domain. For this reason it is common to denote the generator as the pair \((\mathbb B,\mathcal D(\mathbb B))\).

To determine the generator \(\mathbb B\) we can proceed by either differentiation of the semigroup, 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(\mathbb T(t)f-I),\]
or integration;
\[\int_{t=0}^\infty e^{-st}\mathbb T(t)\wrt t = (sI-\mathbb B)^{-1}=:R_s.\]
The operator \(R_s\) is known as the resolvent.

One of the fundamental results of semigroup theory is the Hille-Yosida Theorem which essentially states that a semigroup is entirely characterised by its generator. This is not immediately obvious as the generator is defined on a subset of \(L\) only, whereas \(\{\mathbb T(t)\}\) is defined on all of \(L\).

% \section{Markov processes and semigroups}
% Semigroups arise naturally in Markov processes. Let \(\{X(t):t\geq 0\}\) be a Markov process with state space \(\mathcal S\). The Markov property states that for \(s,t\geq 0\), \(X(t+s)\) is independent of \(X(u),\, 0\leq u<t\) given \(X(t)\). That is, \(P(X(t+s)\in\mathcal{E} \mid X(t), X(u), 0\leq u<t)=P(X(t+s)\in\mathcal{E} \mid X(t))\). If \(P(X(t+s)\in\mathcal{E} \mid X(t))=P(X(s)\in\mathcal{E} \mid X(0))\), that is the distributions are invariant under translation by \(t\), then \(X(t)\) is said to be time-homogeneous. Where necessary, we will use the notation \(X_\mu(t)\) to denote the Markov process with initial distribution \(\mu\), i.e.~\(P(X(0)\in\cdot)=\mu(\cdot)\), and \(X_x(t)\) to be the Markov process with initial distribution \(\delta(x)\). Markov processes can be characterised by \emph{probability kernels}.

% Let \((U,\mathcal U)\) and \((V,\mathcal V)\) be two measure spaces, then a probability kernel \(\nu\), is a mapping \(\nu:U\times \mathcal V\to \mathbb R_+\) such that \(\nu(u,\mathcal{E})\) is \(\mathcal U\)-measurable in \(u\in U\) for fixed \(\mathcal{E}\in \mathcal V\), and a probability measure in \(\mathcal{E}\in \mathcal V\) for fixed \(u\in U\). We say that \(\nu\) is a probability kernel from \(U\) to \(V\).

% With a probability kernel comes associated integral operators. For a measure \(\mu\) on \((U,\mathcal U)\), then 
% \[\mu \nu (\mathcal{E})=\int_{U} \mu(\wrt x) \nu(x,\mathcal{E})\]
% is also a measure on \((U,\mathcal U)\). For a function \(f:V\to \mathbb R\) then 
% \[\nu f (x)=\int_{V}  \nu(x,\wrt y)f(y).\] 
% We can also compose two transition kernels. Let \(\nu\) be a transition kernel from \(U\) to \(V\) and \(\eta\) be a transition kernel from \(V\) to \(W\). Then \(\nu\eta\) is also a probability kernel from \(U\) to \(W\);
% \[\nu\eta(x,\mathcal{E}) = \int_{V}\nu(x,\wrt y)\eta(y,\mathcal{E}).\]

% Let \(\mu_{t}(\cdot)=P(X(t)\in \cdot)\) for \(t\geq 0\) be a one-parameter family of measures on \((\mathcal S,\mathcal F)\), and \(\nu_{t}(x,\cdot)=P(X(t)\in\cdot \mid X(0)=x)\) for \(t\geq 0\) be a one-parameter family of probability kernels from \(\mathcal S\) to \(\mathcal S\). Then \(\mu_t\) and \(\nu_{t}\) characterise the time-homogeneous Markov process \(X(t)\). The probability kernel \(\nu_t\) is referred to as a transition kernel. The kernels \(\nu_t\) form a semigroup since, using the law of total probability and time-homogeneity, 
% \begin{align*}
% \nu_{t+s}(x,\mathcal{E})&=P(X(t+s)\in\mathcal{E}\mid X(0)=x)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(t+s)\in\mathcal{E}\mid X(t)=y)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(s)\in\mathcal{E}\mid X(0)=y)
% \\&=\int_{y\in\mathcal S}\nu_t(x,\wrt y)\nu_s(y,\mathcal{E})=\nu_t\nu_s(x,\mathcal{E}).
% \end{align*}

% We can also construct other semigroups from \(\nu_t\). First we need dome more definitions. Let \((\mathcal S,\mathcal F)\) be a measure space. A \emph{Markov operator on measures} is an operator \(M\) which takes finite measures on \((\mathcal S,\mathcal F)\) to finite measures on \((\mathcal S,\mathcal F)\) such that \(M\) is linear and for a finite measure \(\mu\), \(\mu M(\mathcal S)=\mu (\mathcal S)\). We also have the following. Let \((\mathcal S,\mathcal F, \mu)\) be a measure space and let \(L_1\) be the class of all \(\mu\)-integrable generalised functions on \(\mathcal S\). A linear operator \(P:L_1\to L_1\) is a \emph{Markov operator on densities} when; if \(f\geq 0\) almost everywhere with respect to \(\mu\), then \(fP \geq 0\) almost everywhere with respect to \(\mu\); and if \(f\geq 0\) almost everywhere with respect to \(\mu\), then \(||fP|| = ||f||\). 
% \begin{lem}
% For a Markov operator \(M\) and measure \(\mu <<\lambda\) (absolutely continuous with respect to Lebesgue measure), if \(\mu M <<\lambda\), then \(M\) induces an almost-unique Markov operator on densities \(P\) with respect to \(\mu\). 
% \end{lem}
% Let \((\mathcal S,\mathcal F)\) be a measure space and let \(B(\mathcal S)\) be the class of bounded measurable functions. An operator \(K:B(\mathcal S)\to B(\mathcal S)\) is a \emph{transition operator} when \(K\) is linear and for \(f\), \(\{f_n\}\in B(\mathcal S)\), if \(f\geq 0\) almost everywhere with respect to a measure \(\mu\), \(Kf\geq 0\) almost everywhere with respect to \(\mu\), \(K1(\mathcal S)=1(\mathcal S)\) and if \(f_n\to 0^+\), then \(Kf_n\to 0^+\). 
% \begin{lem}
% Every transition probability kernel \(\nu\) induces a Markov operator on measures, \(\mu M(\mathcal{E}) = \mu \nu(\mathcal{E})\), and a transition operator on functions, \(Kf(x)=\nu f(x)\). In the other direction, every Markov operator induces a probability kernel \(\delta(x)M(\mathcal{E})=\nu(x,\mathcal{E})\), and every transition operator induces a transition kernel \(\nu(x,\mathcal{E})=K1(\mathcal{E})(x)\).
% \end{lem}
